Refactoring the `nls()` function in R

### 2022-05-31 RefactoringNLS/RefactoringNLS.Rmd
add note in acknowledgement -- needs action

An important overall consideration in our work has been the maintainability of
the code base that supports the `nls()` functionality, as we believe that the
existing code makes maintenance and improvement very difficult. Moreover, legacy
applications and examples mean some improvments are blocked. Thus we have not
addressed all the issues raised, but we believe that the discussion and trials
so far are relevant to improving both `nls()` and R generally.

@@

With such a range of features and a long history, it is not surprising that the code has become untidy 
and overly patched. It is, to
our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved. Let
us review some of the issues. We will then propose corrective actions, some of which we have carried out
either in demonstrations or in the distributed coed.

@@

  **Do not use `nls` on artificial "zero-residual" data.**

  The `nls` function uses a relative-offset convergence criterion
  that compares the numerical imprecision at the current parameter
  estimates to the residual sum-of-squares.  This performs well on data of
  the form

  $$  y = f(x, \theta) + eps $$

  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form

  $$  y = f(x, \theta)  $$

  because the criterion amounts to comparing two components of the round-off 
  error.

  If you wish to test `nls` on artificial data please add a noise component, 
  as shown in the example below.

@@

```
t <- -10:10
y <- 100/(1+.1*exp(-0.51*t))
lform<-y~a/(1+b*exp(-c*t))
ldata<-data.frame(t=t, y=y)
plot(t,y)
lstartbad<-c(a=1, b=1, c=1)
lstart2<-c(a=100, b=10, c=1)
nlsr::nlxb(lform, data=ldata, start=lstart2)
nls(lform, data=ldata, start=lstart2, trace=TRUE)
# Fix with scaleOffset
nls(lform, data=ldata, start=lstart2, trace=TRUE, control=list(scaleOffset=1.0))
sessionInfo()
```

@@

```
> t <- -10:10
> y <- 100/(1+.1*exp(-0.51*t))
> lform<-y~a/(1+b*exp(-c*t))
> ldata<-data.frame(t=t, y=y)
> plot(t,y)
> lstartbad<-c(a=1, b=1, c=1)
> lstart2<-c(a=100, b=10, c=1)
> nlsr::nlxb(lform, data=ldata, start=lstart2)
nlsr object: x 
residual sumsquares =  1.007e-19  on  21 observations
    after  13    Jacobian and  19 function evaluations
  name            coeff          SE       tstat      pval      gradient    JSingval   
a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6  
b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3  
c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791  
> nls(lform, data=ldata, start=lstart2, trace=TRUE)
40346.    (1.08e+00): par = (100 10 1)
11622.    (2.93e+00): par = (101.47 0.49449 0.71685)
5638.0    (1.08e+01): par = (102.23 0.38062 0.52792)
642.08    (1.04e+01): par = (102.16 0.22422 0.41935)
97.712    (1.79e+01): par = (100.7 0.14774 0.45239)
22.250    (1.78e+02): par = (99.803 0.093868 0.50492)
0.025789  (1.33e+03): par = (100.01 0.10017 0.50916)
6.0571e-08 (7.96e+05): par = (100 0.1 0.51)
4.7017e-19 (1.86e+04): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
... (approx 40 lines omitted)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) : 
  number of iterations exceeded maximum of 50
> # Fix with scaleOffset
> nls(lform, data=ldata, start=lstart2, trace=TRUE, control=list(scaleOffset=1.0))
40346.      (1.08e+00): par = (100 10 1)
11622.      (2.91e+00): par = (101.47 0.49449 0.71685)
5638.0      (9.23e+00): par = (102.23 0.38062 0.52792)
642.08      (5.17e+00): par = (102.16 0.22422 0.41935)
97.712      (2.31e+00): par = (100.7 0.14774 0.45239)
22.250      (1.11e+00): par = (99.803 0.093868 0.50492)
0.025789    (3.79e-02): par = (100.01 0.10017 0.50916)
6.0571e-08  (5.80e-05): par = (100 0.1 0.51)
4.7017e-19  (1.62e-10): par = (100 0.1 0.51)
Nonlinear regression model
  model: y ~ a/(1 + b * exp(-c * t))
   data: ldata
     a      b      c 
100.00   0.10   0.51 
 residual sum-of-squares: 4.7e-19
Number of iterations to convergence: 8 
Achieved convergence tolerance: 1.62e-10

@@

```{r SESingGrad}
reala=-3
realb=5
realc=0.5
realr=0.7
realm=1
x=1:11 #x values 11 timepoint data
#linear+exponential function
y=reala + realb*realr^(x-realm) + realc*x
testdat=data.frame(x,y)
strt<-list(a=-3, b=5, c=0.5, r=0.7, m=1)
jform<-y~a+b*r^(x-m)+c*x # Formula
linexp=try(nls(jform, data=testdat, start=strt, trace=F))
library(nlsr)
# Note singular values of Jacobian in rightmost column
linexp2<-try(nlxb(jform, data=testdat, start=strt, trace=F))
linexp2
```

@@

`na.action` is an argument to the `nls()` function, but it does not appear obviously in the
source code, often being handled behind the scenes after referencing the option `na.action`.
A useful, but possibly dated, description is given in:
https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/

@@

>>>>

An important overall consideration in our work has been the maintainability of
the code base that supports the `nls()` functionality, as we believe that the
existing code makes maintenance and improvement very difficult. Moreover, legacy
applications and examples mean some improvements are blocked. Thus we have not
addressed all the issues raised, but we believe that the discussion and trials
so far are relevant to improving both `nls()` and R generally.

@@

With such a range of features and a long history, it is not surprising that the code has become untidy 
and overly patched. It is, to
our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved. Let
us review some of the issues. We will then propose corrective actions, some of which we have carried out
either in demonstrations or in the distributed code.

@@

>  **Do not use `nls` on artificial "zero-residual" data.**

>  The `nls` function uses a relative-offset convergence criterion
  that compares the numerical imprecision at the current parameter
  estimates to the residual sum-of-squares.  This performs well on data of
  the form

>  $$  y = f(x, \theta) + eps $$

>  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form

>  $$  y = f(x, \theta)  $$

>  because the criterion amounts to comparing two components of the round-off 
  error.

>  If you wish to test `nls` on artificial data please add a noise component, 
  as shown in the example below.

@@

```
t <- -10:10
y <- 100/(1 + .1 * exp(-0.51 * t))
lform <- y ~ a/(1 + b * exp(-c * t))
ldata <- data.frame(t = t, y = y)
plot(t, y)
lstartbad <- c(a = 1, b = 1, c = 1)
lstart2 <- c(a = 100, b = 10, c = 1)
nlsr::nlxb(lform, data = ldata, start = lstart2)
nls(lform, data = ldata, start = lstart2, trace = TRUE)
# Fix with scaleOffset
nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
sessionInfo()
```

@@

```
> t <- -10:10
> y <- 100/(1 + .1 * exp(-0.51 * t))
> lform <- y ~ a/(1 + b * exp(-c * t))
> ldata <- data.frame(t=t, y=y)
> plot(t,y)
> lstartbad <- c(a = 1, b = 1, c = 1)
> lstart2 <- c(a = 100, b = 10, c = 1)
> nlsr::nlxb(lform, data = ldata, start = lstart2)
nlsr object: x 
residual sumsquares =  1.007e-19  on  21 observations
    after  13    Jacobian and  19 function evaluations
  name            coeff          SE       tstat      pval      gradient    JSingval   
a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6  
b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3  
c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791  
> nls(lform, data = ldata, start = lstart2, trace = TRUE)
40346.    (1.08e+00): par = (100 10 1)
11622.    (2.93e+00): par = (101.47 0.49449 0.71685)
5638.0    (1.08e+01): par = (102.23 0.38062 0.52792)
642.08    (1.04e+01): par = (102.16 0.22422 0.41935)
97.712    (1.79e+01): par = (100.7 0.14774 0.45239)
22.250    (1.78e+02): par = (99.803 0.093868 0.50492)
0.025789  (1.33e+03): par = (100.01 0.10017 0.50916)
6.0571e-08 (7.96e+05): par = (100 0.1 0.51)
4.7017e-19 (1.86e+04): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
... (approx 40 lines omitted)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) : 
  number of iterations exceeded maximum of 50
> # Fix with scaleOffset
> nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
40346.      (1.08e+00): par = (100 10 1)
11622.      (2.91e+00): par = (101.47 0.49449 0.71685)
5638.0      (9.23e+00): par = (102.23 0.38062 0.52792)
642.08      (5.17e+00): par = (102.16 0.22422 0.41935)
97.712      (2.31e+00): par = (100.7 0.14774 0.45239)
22.250      (1.11e+00): par = (99.803 0.093868 0.50492)
0.025789    (3.79e-02): par = (100.01 0.10017 0.50916)
6.0571e-08  (5.80e-05): par = (100 0.1 0.51)
4.7017e-19  (1.62e-10): par = (100 0.1 0.51)
Nonlinear regression model
  model: y ~ a/(1 + b * exp(-c * t))
   data: ldata
     a      b      c 
100.00   0.10   0.51 
 residual sum-of-squares: 4.7e-19
Number of iterations to convergence: 8 
Achieved convergence tolerance: 1.62e-10

@@

```{r SESingGrad}
reala = -3
realb = 5
realc = 0.5
realr = 0.7
realm = 1
x = 1:11 #x values; 11 timepoint data
#linear+exponential function
y=reala + realb * realr^(x - realm) + realc * x
testdat = data.frame(x, y)
strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1)
jform <- y ~ a + b * r^(x - m) + c * x # Formula
linexp=try(nls(jform, data = testdat, start = strt, trace = F))
library(nlsr)
# Note singular values of Jacobian in rightmost column
linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
linexp2
```

@@

`na.action` is an argument to the `nls()` function, but it does not appear obviously in the
source code, often being handled behind the scenes after referencing the option `na.action`.
A useful, but possibly dated, description is given in:
https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/.

@@

??Heather Turner ... (co-mentor ?? how to word?)
